Tools for identifying unexpectedly low microfossil count sums

### 2020-04-07 Rmd/count_check_MS.Rmd
more keywords and funding
reviewers comments

**Keywords:** Microfossil counts; quantitative methods

@@

In general, it is not possible to determine the sum from which percent were calculated, but the properties of community and assemblage counts make it possible to estimate it.
Given that we expect the rarest taxon to be a singleton, the count sum _N_ can be estimated as $1/p_{min} \times 100$ where _p~min~_ is the percent abundance of the rarest taxon.
This method will fail for assemblages without singletons.
We also expect the percent _p_ to be calculated from integer counts, therefore there should be a count sum _N_ such that $p_i/100\times N$ is, within rounding error, an integer for all _i_ taxa.
Possible values for _N_ can be found by a direct search algorithm over the range of plausible values of _N_.
An infinite number of possible count sums that are consistent with the percent exist, but the lowest will give the correct value of _N_ except in cases where the greatest common divisor is greater than one.
These tests are closely related to GRIM [@Brown2017], as all rely on the granularity of percentages calculated from integer data.

@@

Publicly available datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea and other sources.
A range of ecological and palaeoecological data were sought to allow for differences in typical count sums and species richness.
Fossil pollen assemblage in the Neotoma database [@neotoma_2018] (downloaded `r readd(pollen_download_date)`) with small count sums (< 50), that appeared to be percentages, or were documented as back-transformed from digitised data were excluded. 
Datasets with percent data where the minimum count sum was not reported in the associated paper were excluded.
Datasets discussed by Telford [-@Telford2019] as possibly having under-reported count sums were also excluded to avoid double reporting.
Datasets with possible misreporting are anonymised, but no attempt is made to diagnose whether errors or misconduct are responsible.
This paper does not attempt to be an exhaustive survey of all the data available.

@@

Some of the pollen datasets in the Neotoma database have an inexplicably high proportion of assemblages with a greatest common divisor greater than one. 
Here I focus on four datasets produced by the same research group. 
All the datasets, which are relatively recent, include information on the spike used to calculate pollen concentration, so cannot have been digitised.
The four datasets have some assemblages with a greatest common divisor of two interspersed, sometimes regularly, amongst assemblages with the expected greatest common divisor of one.
The assemblages with a greatest common divisor of one typically have several singletons (Table \@ref(tab:neotoma-table)).

@@

My search through archived microfossil data found several datasets where count sums are, or appear to be, smaller than that reported.
Some of the count sums appear to be an order of magnitude below what was reported.
The non-pollen datasets examined here are a convenience sample of available data that met the inclusion criteria.
As such, this analysis cannot be used to accurately determine the prevalence of datasets with possible undercounts, but it appears that a non-negligible fraction of the literature is affected.
It is possible in some cases that the assemblages with low counts were omitted or merged prior to analysis, but the number of assemblages reported was not updated.
Care was taken to identify any such cases by, for example, examining stratigraphic diagrams to see if the low count assemblages were included.
In some cases, the assemblages with low apparent counts can be seen in the published stratigraphic diagrams.

@@

Some data were obtained from the Neotoma Paleoecology Database (http://www.neotomadb.org), and the work of the data contributors and the Neotoma community is gratefully acknowledged.

>>>>

**Keywords:** Microfossils assemblages, Counts; Quantitative methods, Reproducible research, Meta-research

@@

In general, it is not possible to determine the sum from which percentages were calculated, but the properties of community and assemblage counts make it possible to estimate it.
Given that we expect the rarest taxon to be a singleton, the count sum _N_ can be estimated as $1/p_{min} \times 100$ where _p~min~_ is the percent abundance of the rarest taxon.
This method will fail for assemblages without singletons.
We also expect the percent _p_ to be calculated from integer counts, therefore there should be a count sum _N_ such that $p_i/100\times N$ is, within rounding error, an integer for all _i_ taxa.
Possible values for _N_ can be found by a direct search algorithm over the range of plausible values of _N_.
An infinite number of possible count sums that are consistent with the percent exist, but the lowest will give the correct value of _N_ except in cases where the greatest common divisor is greater than one.
These tests are closely related to GRIM [@Brown2017], as all rely on the granularity of percentages calculated from integer data.

@@

Publicly available datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea, and other sources.
A range of ecological and palaeoecological data were sought to allow for differences in typical count sums and species richness.
Fossil pollen assemblage in the Neotoma database [@neotoma_2018] (downloaded `r readd(pollen_download_date)`) with small count sums (< 50), that appeared to be percentages, or were documented as back-transformed from digitised data were excluded. 
Datasets with percent data where the minimum count sum was not reported in the associated paper were excluded.
Datasets discussed by Telford [-@Telford2019] as possibly having misreported the minimum count sum were also excluded to avoid double reporting.
Datasets with possible misreporting are anonymised, but no attempt is made to diagnose whether errors or misconduct are responsible.
This paper does not attempt to be an exhaustive survey of all the data available.

@@

Some of the pollen datasets in the Neotoma database have an inexplicably high proportion of assemblages with a greatest common divisor greater than one. 
Here I focus on four datasets produced by one research group. 
All the datasets, which are relatively recent, include information on the spike used to calculate pollen concentration, so cannot have been digitised.
The four datasets have some assemblages with a greatest common divisor of two interspersed, sometimes regularly, amongst assemblages with the expected greatest common divisor of one.
The assemblages with a greatest common divisor of one typically have several singletons (Table \@ref(tab:neotoma-table)).

@@

My search through archived microfossil data found several datasets where count sums are, or appear to be, smaller than that reported.
Some of the count sums appear to be an order of magnitude below what was reported.
The non-pollen datasets examined here are a convenience sample of available data that met the inclusion criteria.
As such, this analysis cannot be used to accurately determine the prevalence of datasets with possible undercounts, but it appears that a small but non-negligible fraction of the literature is affected.
It is possible in some cases that the assemblages with low counts were omitted or merged prior to analysis, but the number of assemblages reported was not updated.
Care was taken to identify any such cases by, for example, examining stratigraphic diagrams to see if the low count assemblages were included.
In some cases, the assemblages with low apparent counts can be seen in the published stratigraphic diagrams.

@@

Some data were obtained from the Neotoma Paleoecology Database (http://www.neotomadb.org), and the work of the data contributors and the Neotoma community is gratefully acknowledged. This work was partly supported by Norwegian Research Council project PalaeoDrivers (213607).
